DeepPicar: A Low-cost Deep Neural Network-based Autonomous Car

### 2018-04-20 paper/background.tex
Changes to sections I through III

Using deep neural networks involves two distinct
phases~\cite{NVIDIA2015}. The first
phase is \emph{training} during which the weights of the network are
incrementally updated by backpropagating errors it sees from the
training examples. Once the network is trained---i.e., the weights of
the network minimize errors in the training examples---the next phase
is \emph{inferencing}, during which unseen data is fed to the network
as input to produce predicted output (e.g., predicted image
classification). In general, the training phase is more computationally
intensive and requires high throughput, which is generally not
available on embedded platforms. The inferencing phase, on the
other hand, is relatively less computationally intensive and latency becomes
as important, if not moreso, as computational throughput, because many
use cases have strict real-time requirements.

@@

Accelerating AI workloads, especially inferencing
operations, has received a lot of attention from academia and industry
in recent years as applications of deep learning are broadening to include
areas of real-time embedded systems such as autonomous vehicles. These
efforts include the development of various heterogeneous architecture-based 
system-on-a-chips (SoCs) that may include multiple cores, GPU,
DSP, FPGA, and neural network optimized ASIC hardware~\cite{Jouppi2017}.
Consolidating multiple tasks on SoCs with a lot of shared hardware
resources while guaranteeing real-time performance is also an active
research area, which is orthogonal to improving raw
performance. Consolidation is necessary for efficiency, but unmanaged 
interference can nullify the benefits of consolidation~\cite{Kim2016}.
%% For these reasons, finding a good computing platform is a
%% non-trivial task, one that requires a deep understanding of the
%% workloads and the hardware platform being utilized.

>>>>

Using deep neural networks involves two distinct
phases~\cite{NVIDIA2015}. The first
phase is \emph{training}, during which the weights of the network are
incrementally updated by backpropagating errors it sees from the
training examples. Once the network is trained---i.e., the weights of
the network minimize errors in the training examples---the next phase
is \emph{inferencing}, during which unseen data is fed to the network
as input to produce predicted output (e.g., predicted image
classification). In general, the training phase is more computationally
intensive and requires high throughput, which is generally not
available on embedded platforms. The inferencing phase, on the
other hand, is relatively less computationally intensive and latency becomes
as important, if not moreso, as computational throughput, because many
use cases have strict real-time requirements.

@@

Accelerating AI workloads, especially inferencing
operations, has received a lot of attention from academia and industry
in recent years as applications of deep learning are broadening to include
areas of real-time embedded systems such as autonomous vehicles. These
efforts include the development of various heterogeneous architecture-based 
system-on-a-chip (SoCs) that may include multiple cores, GPU,
DSP, FPGA, and neural network optimized ASIC hardware~\cite{Jouppi2017}.
Consolidating multiple tasks on SoCs with a lot of shared hardware
resources while guaranteeing real-time performance is also an active
research area, which is orthogonal to improving raw
performance. Consolidation is necessary for efficiency, but unmanaged 
interference can nullify the benefits of consolidation~\cite{Kim2016}.
%% For these reasons, finding a good computing platform is a
%% non-trivial task, one that requires a deep understanding of the
%% workloads and the hardware platform being utilized.

### 2018-04-20 paper/intro.tex
Changes to sections I through III

% big problem
Executing these AI workloads on an embedded computing platform 
poses several additional challenges. First, many AI workloads,
especially those in vehicles, are computationally demanding and have
strict real-time requirements.
For example, computing latency in a vision-based object
detection task may be directly linked to the safety of the vehicle. This
requires a high computing capacity as well as the means to guaranteeing
the timings. On the other hand, the computing hardware platform must
also satisfy cost, size, weight, and power constraints, which require a
highly efficient computing platform. These two conflicting
requirements complicate the platform selection process as observed in
~\cite{Otterness2017}.

@@

% Findings:
%
% - we find DNN processing is an ideal real-time workload from the
%   perspective of timing predictability as the amount of computation
%   needed is fixed at design time and doesn't change over input.
%
% - real-time processing of non-trivial DNN inferencing is feasible
%   on today's embedded multicore platform, even as inexpensive as
%   raspberry pi3.
%
% - all tested platforms have enough computing capacity to consolidate 
%   multiple DeepPicar's CNN workloads or other compute intensive tasks
%   simultaneously.
%
% - however, when consolidating multiple workloads shared resource
%   contention is an important issue. we carefully analyzed
%   sensitivity of DeepPicar's workload with respect to shared
%   hardware resources.
%
% - in particular, we find DNN processing is highly sensitive to
%   memory performance. performance severely affected when memory
%   bandwidth is contented.
%
% - however, we find that dnn processing is not sensitive to cache
%   space. we find applying cache partitioning does not help improve
%   isolation of DNN workload when memory bandwidth is contended.
Our main observations are as follows.
First, we find that real-time processing of the CNN inferencing is
feasible on contemporary embedded computing platforms, even one   
as inexpensive as the Raspberry Pi 3. Second, while consolidating
additional CNN models and tasks on a single multicore platform is
feasible but the impact of shared resource contention can be
catastrophic---we observe up to 11.6X slowdown even when the CNN model
was given a dedicated core.
Third, our evaluation of existing cache
partitioning and memory bandwidth throttling
techniques~\cite{yun2014rtas,Yun2013} shows that cache partitioning is
not effective in protecting the CNN inferencing task while memory
bandwidth throttling is quite effective.

@@

The remainder of the paper is organized as follows.
Section~\ref{sec:background} provides a background on the application
of nueral networks in autonomous driving.
Section~\ref{sec:overview} gives an overview of the DeepPicar
testbed. Section~\ref{sec:evaluation} presents extensive real-time
performance evaluation results we collected on the testbed.
Section~\ref{sec:comparison} offers a comparison between  
the Raspberry Pi 3 and other embedded computing platforms.
%% We discuss more detailed training methods and system-level issues
%% in~\ref{sec:discussion}. 
We review related work in
Section~\ref{sec:related} and conclude in
Section~\ref{sec:conclusion}.

>>>>

% big problem
Executing these AI workloads on an embedded computing platform 
poses several additional challenges. First, many AI workloads,
especially those in vehicles, are computationally demanding and have
strict real-time requirements.
For example, computing latency in a vision-based object
detection task may be directly linked to the safety of the vehicle. This
requires a high computing capacity as well as the means to guaranteeing
the timings. On the other hand, the computing hardware platform must
also satisfy cost, size, weight, and power constraints, which require a
highly efficient computing platform. These two conflicting
requirements complicate the platform selection process as observed 
in~\cite{Otterness2017}.

@@

% Findings:
%
% - we find DNN processing is an ideal real-time workload from the
%   perspective of timing predictability as the amount of computation
%   needed is fixed at design time and doesn't change over input.
%
% - real-time processing of non-trivial DNN inferencing is feasible
%   on today's embedded multicore platform, even as inexpensive as
%   raspberry pi3.
%
% - all tested platforms have enough computing capacity to consolidate 
%   multiple DeepPicar's CNN workloads or other compute intensive tasks
%   simultaneously.
%
% - however, when consolidating multiple workloads shared resource
%   contention is an important issue. we carefully analyzed
%   sensitivity of DeepPicar's workload with respect to shared
%   hardware resources.
%
% - in particular, we find DNN processing is highly sensitive to
%   memory performance. performance severely affected when memory
%   bandwidth is contented.
%
% - however, we find that dnn processing is not sensitive to cache
%   space. we find applying cache partitioning does not help improve
%   isolation of DNN workload when memory bandwidth is contended.
Our main observations are as follows.
First, we find that real-time processing of the CNN inferencing is
feasible on contemporary embedded computing platforms, even one   
as inexpensive as the Raspberry Pi 3. Second, while consolidating
additional CNN models and tasks on a single multicore platform is
feasible, the impact of shared resource contention can be
catastrophic---we observe up to 11.6X slowdown even when the CNN model
was given a dedicated core.
Third, our evaluation of existing cache
partitioning and memory bandwidth throttling
techniques~\cite{yun2014rtas,Yun2013} shows that cache partitioning is
not effective in protecting the CNN inferencing task while memory
bandwidth throttling is quite effective.

@@

The remainder of the paper is organized as follows.
Section~\ref{sec:background} provides a background on the application
of neural networks in autonomous driving.
Section~\ref{sec:overview} gives an overview of the DeepPicar
testbed. Section~\ref{sec:evaluation} presents extensive real-time
performance evaluation results we collected on the testbed.
Section~\ref{sec:comparison} offers a comparison between  
the Raspberry Pi 3 and other embedded computing platforms.
%% We discuss more detailed training methods and system-level issues
%% in~\ref{sec:discussion}. 
We review related work in
Section~\ref{sec:related} and conclude in
Section~\ref{sec:conclusion}.

### 2018-04-20 paper/overview.tex
intro revision
Minor changes
Changes to sections I through III

\section{DeepPicar Overview}\label{sec:overview}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=.38\textwidth]{figs/DeepPicar_platform}
  \caption{DeepPicar platform.}
  \label{fig:overview}
\end{figure}

@@

% inferencing on pi3
Once the network is trained on the desktop computer, the trained model
is copied back to the Raspberry Pi 3. The network is then used
by the car's main controlller, which feeds image frames from the web
camera as input to the network in real-time. At each control period,
the network produced steering angle output is converted into the PWM values
of the car's steering motor. Figure~\ref{fig:controlloop} shows simplified 
pseudo code of the controller's main loop. Among the five steps, the 3rd step, 
network inferencing, is the most computationally intensive and dominates the
execution time.

@@

Another factors that can potentially affect the prediction accuracy of
the CNN, are camera and actuator (motor) control latencies. The camera
latency is from the time the camera sensor observes the scene to the
time the computer actually reads the digitized image data. This time
can be noticable depending on the camera used and the data processing
time of the computing platform. Higher camera latency could
negatively affect control performance, because the CNN would analyze
stale scenes. The actuator (motor) control latency from the time
the control output is sent to the steering motor to the time the motor
actually moved at a desired position, which also can takes
considerable time. In our platform, the combined latency is measured
to be around 50 ms, which is reasonble.
%% In other words, %% in the recorded video and control data, 
%% a control action of the CNN appears to be applied 50 ms later.
%% Considering that camera's framerate is 30Hz
%% (33.3 ms/frame), this is about two frames after the control action.
%% We experimentally measured the camera
%% latency and found it to be around 50-100 ms.
If this value is too high, control performance may suffer.
Our initial prototype suffered this problem as the observed latency
was as high as 300 ms, which negatively affected control performance.
For reference, the latency of human perception is known to be as fast
as 13 ms~\cite{ThomasBurger2015}. 
% https://www.pubnub.com/blog/2015-02-09-how-fast-is-realtime-human-perception-and-technology/

Our trained CNN models showed good prediction accuracy, successfully
navigating several different tracks we used for training.
For instance, the DeepPicar could remain on Track 1
(Figure~\ref{fig:track}) over 10 minutes at a moderate speed (50\%
throttle), at which point we stopped the experiment, and more than one
minute at a higher speed (75\% throttle)~\footnote{Self-driving videos: \url{https://photos.app.goo.gl/q40QFieD5iI9yXU42}
  %% \url{https://photos.app.goo.gl/ce93sU7jPk4ywO8u2}
}. Running at
higher speed is inherently more challenging because the CNN controller
has less time to recover from mistakes (bad predictions).  Also, we
find that the prediction accuracy is significantly affected by the
quality of training data as well as various environmental factors such
as lighting conditions. We plan to investigate more systematic ways
to improve the CNN's prediction accuracies.

>>>>

\section{DeepPicar}\label{sec:overview}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=.4\textwidth]{figs/DeepPicar_platform}
  \caption{DeepPicar platform.}
  \label{fig:overview}
\end{figure}

@@

% inferencing on pi3
Once the network is trained on the desktop computer, the trained model
is copied back to the Raspberry Pi 3. The network is then used
by the car's main controlller, which feeds image frames from the web
camera as input to the network in real-time. At each control period,
the network produced steering angle output is converted into the PWM values
of the car's steering motor. Figure~\ref{fig:controlloop} shows simplified 
pseudo code of the controller's main loop. Among the five steps, the 3rd step, 
network inferencing, is the most computationally intensive and is expected to 
dominate the execution time.

@@

Other factors that can potentially affect the prediction accuracy of
the CNN, are camera and actuator (motor) control latencies. The camera
latency is defined as the time the camera sensor observes the scene to the
time the computer actually reads the digitized image data. This time
can be noticable depending on the camera used and the data processing
time of the computing platform. Higher camera latency could
negatively affect control performance, because the CNN would analyze
stale scenes. The actuator (motor) control latency from the time
the control output is sent to the steering motor to the time the motor
actually moved at a desired position, which also can takes
considerable time. In our platform, the combined latency is measured
to be around 50 ms, which is reasonble.
%% In other words, %% in the recorded video and control data, 
%% a control action of the CNN appears to be applied 50 ms later.
%% Considering that camera's framerate is 30Hz
%% (33.3 ms/frame), this is about two frames after the control action.
%% We experimentally measured the camera
%% latency and found it to be around 50-100 ms.
If this value is too high, control performance may suffer.
Our initial prototype suffered this problem as the observed latency
was as high as 300 ms, which negatively affected control performance.
For reference, the latency of human perception is known to be as fast
as 13 ms~\cite{ThomasBurger2015}. 
% https://www.pubnub.com/blog/2015-02-09-how-fast-is-realtime-human-perception-and-technology/

Our trained CNN models showed good prediction accuracy, successfully
navigating several different tracks we used for training.
For instance, the DeepPicar could remain on Track 1
(Figure~\ref{fig:track}) for over 10 minutes at a moderate speed (50\%
throttle), at which point we stopped the experiment, and more than one
minute at a higher speed (75\% throttle)~\footnote{Self-driving videos: \url{https://photos.app.goo.gl/q40QFieD5iI9yXU42}
  %% \url{https://photos.app.goo.gl/ce93sU7jPk4ywO8u2}
}. Running at
higher speed is inherently more challenging because the CNN controller
has less time to recover from mistakes (bad predictions).  Also, we
find that the prediction accuracy is significantly affected by the
quality of training data as well as various environmental factors such
as lighting conditions. We plan to investigate more systematic ways
to improve the CNN's prediction accuracies.
